---
title: "Statistical Analysis for Capstone Project"
author: "Laura Rose"
date: "June 26, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
  
  Prior to applying machine learning techniques to the data, it is important to look at some basic statistics and visualizations to better understand the data. This also helps one to be aware of any potential problems that could be encountered in applying machine learning.
  
  The first step taken to understand the data was to summarize all variables. Results are found in the table below. Data is in a weekly format, beginning in October 2013 (week 40 of 2013) and ending in March 2017 (week 13 of 2017). Weekly_shipments is the amount of a pharmaceutical ingredient of Tylenol requested by date by the company's customer. The request date may not necessarily correspond to ship date, but rather it is the date which the customer demanded the product. In this sense, request date is a better measure of the timing of true demand instead of ship date, which may be altered due to backorder, etc.
  
  Total.Flu is the total number of specimens tested by clinical and public health laboratories reported by the CDC. Prior to the 2015-2016 season (beginning the 40th week of 2015), the data from both types of laboratory were combined. Beginning in the 2015-2016 season, the data were manually combined from each type of laboratory to form a variable comparable to earlier data. Total.ILI is the total number of cases of Influenza-like Illness symptoms reported by sentinel providers by the CDC. Positive.Specimens is the number of specimens that tested positive for any of the strains of flu considered. This was calculated by multiplying the percent positive by the total number tested.
  
  Weekly_tylenol_vec is units of Tylenol sold into the channels (retail pharmacies, hospitals, clinics, etc.) reported by QuintilesIMS, a global information company directed at the healthcare industry. Sales into the channels are not the same as sales to consumers. Unfortunately, this Tylenol data was not available in weekly format, so the weekly data was calculated by dividing each month by four or five and using this as the weekly amount for each month. February, May, August, and November were chosen to be five-week months based on my company's dating convention. December 2014 was chosen to be a five-week month to align with the CDC data, which listed 2014 as a 53-week month. Of note is the large number of 0's in the dependent variable weekly_shipments, as shown in the table below. These 0s represent weeks in which the customer did not request any product, so it makes sense to keep them in the data.
  
  It is also interesting to note the similar pattern of distribution for Total.Flu and Total.ILI. This is reasonable because potential flu specimens and Influenze-like Illness cases are presumably highly correlated. The median for Total.ILI is slightly higher than the median for Total.Flu, but the maximum of Total.Flu is about twenty-one percent higher than the maximum for Total.ILI.
  
```{r}
summary(final_project_data3_df)
```

  A line chart comparing Total.ILI, Total.Flu, and Positive.Specimens is shown below. It is noteworthy that the peak of each flu season occurred around January in 2014 and 2015, but was slightly later in 2016 and 2017. Also, 2017 seems to have been particularly strong year for the flu in terms of specimens tested, but not in terms of the number of positive specimens. 
  
  Regarding the troughs for each year, these generally occur around July, which makes sense since this is the middle of the summer in the U.S.

```{r}
project_data_plot3 <- final_project_data3 %>% gather(key, value, Total.Flu, Total.ILI, Positive.Specimens)
```

```{r}
ggplot(project_data_plot3, aes(x = Date, y = value, col = key)) +
  geom_line()
```

  However, this peak and trough pattern differs somewhat for the weekly_shipments variable, as shown in the graph below. Note that although the timing of the peaks for weekly_shipments and the Total.Flu and Total.ILI variables approximately match in 2016 and 2017, there are also instances of a high value of weekly_shipments with a low value of Flu/ILI. For example, the Flu/ILI trough of mid-2016 corresponds to relatively high shipments volume. Also, in 2015, the highest value of shipments occurred after the peak at the beginning of the year.
  
  Per the VP of Active Pharmaceutical Ingredient (API) Sales at the company, these periods of high demand outside of the flu season may be due to the fact that pharmacies restock in July in preparation for the upcoming cold/flu season. This would drive increased demand in weekly_shipments since the manufacturer would need more ingredients for Tylenol to replenish depleted finished good inventories.

```{r}
project_data_plot4 <- final_project_data3 %>% gather(key, value, Total.Flu, Total.ILI, Positive.Specimens, weekly_shipments)
ggplot(project_data_plot4, aes(x = Date, y = value, col = key)) +
  geom_line()
```

  To further illustrate this point, a Pearson correlation matrix is shown below. Note that all the pairwise correlations between Total.ILI, Total.Flu, and Positive.Specimens are greater than 0.9, which suggest that multicollinearity could be present if more than one are included in the model. Unfortunately, the correlations between all of the explanatory variables and the dependent variable weekly_shipments are quite low. This raises the question of whether these variables are truly explanatory in regards to the weekly_shipments.


```{r}

cor(final_project_data3_df, method = "pearson")
```

  The question of whether the explanatory variables actually explain any variation in the dependent variable is visually illustrated below with scatterplots. At first glance, it appears that the relationship depicted on these scatterplots could best be represented with an almost horizontal line, implying little correlation.

```{r}
ggplot(final_project_data3, aes(x = weekly_tylenol_vec, y = weekly_shipments)) +
  geom_point()
```
```{r}
ggplot(final_project_data3, aes(x = Total.Flu, y = weekly_shipments)) +
  geom_point()
```

```{r}
ggplot(final_project_data3, aes(x = Total.ILI, y = weekly_shipments)) +
  geom_point()
```

```{r}
ggplot(final_project_data3, aes(x = Positive.Specimens, y = weekly_shipments)) +
  geom_point()
```


Finally, below are histograms of the dependent and independent variables. The variables do not appear to be normally distributed and seem skewed toward their lower values. This should not be problematic for linear regression since that method does not require an assumption of normality in the variables, but rather that the errors are independently and identically distributed.

```{r}
ggplot(final_project_data3, aes(x = weekly_shipments)) +
  geom_histogram(binwidth = 5000)
```

```{r}
ggplot(final_project_data3, aes(x = Total.Flu)) +
  geom_histogram(binwidth = 1000)
```

```{r}
ggplot(final_project_data3, aes(x = Total.ILI)) +
  geom_histogram(binwidth = 750)
```

```{r}
ggplot(final_project_data3, aes(x = weekly_tylenol_vec)) +
  geom_histogram(binwidth = 750000)
```

```{r}
ggplot(final_project_data3, aes(x = Positive.Specimens)) +
  geom_histogram(binwidth = 500)
```